Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Tumors

### 2018-06-14 submission/manuscript.Rmd
spell check and add nick changes to abstract

**Background.** Colorectal cancer (CRC) is a growing health concern with a large percentage of the risk of developing disease being due to environmental factors. Certain bacterial community members have been associated to increase CRC while other taxa have been associated with normal colons. Some of the taxa associated with normal colons can use fiber to produce short-chain fatty acids (SCFAs) that can inhibit tumor growth in model systems. However, the data supporting the importance of SCFAs in human CRC is less certain. Here, we test the hypothesis that SCFA concentration changes in individuals with colorectal tumors.

**Methods.** We analyzed a cross-sectional (n=`r length(metai$sample)`) and pre- and post-treatment (n=`r length(metaf$initial)`) group for their concentrations of acetate, butyrate, and propionate. Analysis also included tumor classification models using Random Forest, imputed gene relative abundance with PICRUSt, and metagenomic sequencing on a subset (n=`r length(geof_samples$X30)`) of the total cross-sectional group.

**Results.** No difference in SCFA concentration was found between normal individuals and patients with adenoma or carcinoma (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15). There was no difference in classification models with or without SCFAs in their ability to predict adenoma or carcinoma (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05). Using metagenomic sequencing, there was also no difference in genes involved with SCFA synthesis (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70).

**Conclusions.** Although our data does not support the hypothesis that SCFAs decrease in individuals that have colorectal tumors, there may be specific scenarios where SCFAs may still be beneficial for treatment of CRC. Alternatively, there may be other metabolites that have not been thoroughly investigated that are more important to the development of human CRC.

@@

Colorectal cancer (CRC) is currently the third leading cancer-related cause of death within the US [@Haggar2009; @Siegel2016]. Although there is a genetic component to the disease, the envrionment has been attributed as an important and large risk factor for CRC [@Lichtenstein2000]. This has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [@Zackular2013; @Baxter2014; @Zackular2015] and whether the bacterial community is altered [@Shah2017; @meta_analysis_crc_Sze2018]. Multiple reports in case/control studies on the microbiota and CRC have identified taxa commonly associated with the normal colons to be decreased in patients with carcinoma tumors [@Zeller2014; @Baxter2016; @Kostic2011]. Many of these taxa actively produce short-chain fatty acids (SCFAs) from fiber that are a part of our general diet [@Holscher2017]. The most extensively studied of these SCFAs are acetate, butyrate, and propionate [@Louis2016]. Overall, CRC is a growing health problem and the microbiota through SCFAs have shown promise as a potential treatment option.

@@

To accomplish this task we will directly measure the concentration of acetate, butyrate, and propionate within fecal samples for two different groups. The first group had a sample obtained at a single cross sectional point in time while the second group had samples obtained before (pre-) and after (post-)treatment for colorectal tumors. Additionally, we also assessed whether adding SCFA concentrations to OTU data improves the classification of adenoma or carcinoma using the Random Forest algorithm [@randomforest_citation_2002], used PICRUSt [@Langille2013] and metagenomic sequencing to assess the genes involved in SCFA synthesis, and analyzed how well 16S rRNA gene sequencing can predict SCFA concentrations. This  investigation provides additional information as to whether the taxa that are decreased in carcionma use SCFAs to help protect against colon tumorigenesis.

@@

**Changes in genes for enzymes involved in SCFA synthesis are not associated with tumors.** Using a list of specific genes that are important for the synthesis of SCFAs [Table S1], we looked for differences in gene abundance based on individuals having normal colons, adenomas, or carcinomas. First, using imputed gene relative abundance based on 16S rRNA gene sequencing we found no difference in any of the genes inolved with acetate, butyrate, or propionate synthesis (P-value `r ifelse(min(pull(picrust_select, bh)) > 0.90, invisible(">"), invisible("<"))` 0.90) [Table S2]. This similarity between groups is highlighted by visualizing genes important in butyrate synthesis [Figure 3A]. Using a Wilcoxson rank-sum test, there also was no difference in imputed gene relative abundance between pre- and post-treatment samples for any genes involved with SCFA synthesis (P-value `r ifelse(min(pull(picrust_select_treat, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3]. Next, we took a subset of these `r length(metai$sample)` fecal samples (n=`r length(geof_samples$X1)`) and used metagenomic sequencing to confirm these results. Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals with normal colons (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenoma (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinoma (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S4]. This lack of difference is highlighted when we visualize the results for butyrate kinase [Figure 3B]. These observations provide evidence that the gene content does not change due to colorectal tumors. Although these results suggest that SCFA concentrations do not change due to colorectal tumor, there may be some errors present within our results that may account for the lack of associations.

**Expected taxa are associated with higher SCFA concentrations regardless of tumor status.** Using OTU data we built Random Forest models to classify higher than median and lower than median SCFA concentrations. Overall, OTU data had a reasonable ability to classify high and low SCFA concentrations [Figure S1A]. However, these models tended to be overfit, suggesting that rarer taxa may be important for this classification [Figure S1A]. The most important OTUs to these models (assessed with mean decrease in accuracy (MDA)) were to taxa that are normally associated with SCFA production [Figure S1B]. These results highlight that there are no immediately apparent errors in our data because SCFA concentrations are associated with taxa known to produce acetate, butyrate, and propionate. Additionally, OTUs associated with these taxa are the most important to models that can classify high and low SCFA concentrations. Overall, our results are robust and do not support the hypothesis that differences in SCFA concentrations are associated with colorectal cancer.

@@

**Study design and sampling.** The overall protocol has been described in detail previously [@normalization_sze2017; @crc_model_baxter2016]. In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before (pre-) and after (post-) treatment of a patient's tumor (n=`r length(metaf$initial)`). For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from `r min(metaf$time)` - `r max(metaf$time)` days. Our use of treatment has been previously defined as encompassing removal of a tumor with or without chemotherapy and radiation [@normalization_sze2017]. Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@normalization_sze2017; @crc_model_baxter2016]. The University of Michigan Institutional Review Board approved the study and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Decleration.

**Measuring specific SCFAs.** Our protocol for the measurement of acetate, butyrate, and propionate followed a previously published protocol that used a High-Performance Liquid Chromatography (HPLC) machine [@scfa_measures_venkataraman2016]. The following changes to this protocol included the use of frozen fecal samples suspended in 1ml of PBS instead of fecal suspensions in DNA Genotek OmniGut tubes, and the use of the acutal weight of fecal samples instead of the average weight for SCFA concentration normalizations. These methodological changes did not affect the overall median concentrations of these SCFAs between the two studies (see Table 1 [@scfa_measures_venkataraman2016] and Figure 1 here).

@@

**Generating imputed metagenomes.** The use of PICRUSt version 1.1.2 with the recommended standard operating protocol [@Langille2013] was used. Briefly, the mothur shared file and metadata was converted into a biom formated table using the biom convert function, the subsequent biom file was processed with the 'normalize_by_copy_number.py' function, and subsequent imputed metagenomes created using the 'predict_metagenomes.py' function.

@@

The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study. We would also like to thank Kwi Kim and Thomas M Schmidt for their help in running the short-chian fatty acid analysis on the High-Performace Liquid Chromatography machine at the University of Michigan. Salary support for Marc A. Sze came from the Canadian Institute of Health Research and NIH grant UL1TR002240. Salary support for Patrick D. Schloss came from NIH grants P30DK034933 and 1R01CA215574.

@@

**Figure S1. OTU-based Random Forest models of SCFA concentrations.** The train and test results of 100 different 80/20 OTU-based models with 10-fold CV based on higher or lower than the medain SCFA concentration using classification Random Forest (A). The top 10 OTUs based on mean decrease in accuracy (MDA) for each model, colored by their lowest taxonomic identification (B).

>>>>

**Background.** Colorectal cancer (CRC) is a growing health concern with the majority of the risk for developing disease being due to environmental factors. The microbiota is one of these environmental factors with certain bacterial community members being associated with CRC, while other taxa are associated to colons without tumors. Some of the taxa associated to colons without tumors can use fiber to produce short-chain fatty acids (SCFAs) that can inhibit tumor growth in model systems. However, the data supporting the importance of SCFAs in human CRC is less certain. Here, we test the hypothesis that SCFA concentrations are different in individuals with colorectal tumors.

**Methods.** We analyzed a cross-sectional (n=`r length(metai$sample)`) and longitudinal pre- and post-treatment (n=`r length(metaf$initial)`) group for their concentrations of acetate, butyrate, and propionate. Analysis also included tumor classification models using Random Forest, imputed gene relative abundance with PICRUSt, and metagenomic sequencing on a subset (n=`r length(geof_samples$X30)`) of the total cross-sectional group.

**Results.** No difference in SCFA concentrations were found between individuals without tumors and patients with adenomas or carcinomas (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15). There was no difference in classification models with or without SCFAs in their ability to predict patients with adenomas or carcinomas versus individuals without tumors (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05). Using metagenomic sequencing, there was also no difference in genes involved with SCFA synthesis between individuals without tumors and patients with adenomas or carcinomas (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70).

**Conclusions.** Although our data does not support the hypothesis that SCFAs are different in individuals that have colorectal tumors, there may be context specific scenarios where SCFAs may still be beneficial for treatment of CRC. Alternatively, there may be other mechanisms that have not been thoroughly investigated that are more important to the development of human CRC.

@@

Colorectal cancer (CRC) is currently the third leading cancer-related cause of death within the US [@Haggar2009; @Siegel2016]. Although there is a genetic component to the disease, the environment has been attributed as an important and large risk factor for CRC [@Lichtenstein2000]. This has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [@Zackular2013; @Baxter2014; @Zackular2015] and whether the bacterial community is altered [@Shah2017; @meta_analysis_crc_Sze2018]. Multiple reports in case/control studies on the microbiota and CRC have identified taxa commonly associated with the normal colons to be decreased in patients with carcinoma tumors [@Zeller2014; @Baxter2016; @Kostic2011]. Many of these taxa actively produce short-chain fatty acids (SCFAs) from fiber that are a part of our general diet [@Holscher2017]. The most extensively studied of these SCFAs are acetate, butyrate, and propionate [@Louis2016]. Overall, CRC is a growing health problem and the microbiota through SCFAs have shown promise as a potential treatment option.

@@

To accomplish this task we will directly measure the concentration of acetate, butyrate, and propionate within fecal samples for two different groups. The first group had a sample obtained at a single cross sectional point in time while the second group had samples obtained before (pre-) and after (post-)treatment for colorectal tumors. Additionally, we also assessed whether adding SCFA concentrations to OTU data improves the classification of adenoma or carcinoma using the Random Forest algorithm [@randomforest_citation_2002], used PICRUSt [@Langille2013] and metagenomic sequencing to assess the genes involved in SCFA synthesis, and analyzed how well 16S rRNA gene sequencing can predict SCFA concentrations. This  investigation provides additional information as to whether the taxa that are decreased in carcinoma use SCFAs to help protect against colon tumorigenesis.

@@

**Changes in genes for enzymes involved in SCFA synthesis are not associated with tumors.** Using a list of specific genes that are important for the synthesis of SCFAs [Table S1], we looked for differences in gene abundance based on individuals having normal colons, adenomas, or carcinomas. First, using imputed gene relative abundance based on 16S rRNA gene sequencing we found no difference in any of the genes involved with acetate, butyrate, or propionate synthesis (P-value `r ifelse(min(pull(picrust_select, bh)) > 0.90, invisible(">"), invisible("<"))` 0.90) [Table S2]. This similarity between groups is highlighted by visualizing genes important in butyrate synthesis [Figure 3A]. Using a Wilcoxon rank-sum test, there also was no difference in imputed gene relative abundance between pre- and post-treatment samples for any genes involved with SCFA synthesis (P-value `r ifelse(min(pull(picrust_select_treat, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3]. Next, we took a subset of these `r length(metai$sample)` fecal samples (n=`r length(geof_samples$X1)`) and used metagenomic sequencing to confirm these results. Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals with normal colons (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenoma (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinoma (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S4]. This lack of difference is highlighted when we visualize the results for butyrate kinase [Figure 3B]. These observations provide evidence that the gene content does not change due to colorectal tumors. Although these results suggest that SCFA concentrations do not change due to colorectal tumor, there may be some errors present within our results that may account for the lack of associations.

**Expected taxa are associated with higher SCFA concentrations regardless of tumor status.** Using OTU data we built Random Forest models to classify higher than median and lower than median SCFA concentrations. Overall, OTU data had a reasonable ability to classify high and low SCFA concentrations [Figure S1A]. However, these models tended to be over fit, suggesting that rarer taxa may be important for this classification [Figure S1A]. The most important OTUs to these models (assessed with mean decrease in accuracy (MDA)) were to taxa that are normally associated with SCFA production [Figure S1B]. These results highlight that there are no immediately apparent errors in our data because SCFA concentrations are associated with taxa known to produce acetate, butyrate, and propionate. Additionally, OTUs associated with these taxa are the most important to models that can classify high and low SCFA concentrations. Overall, our results are robust and do not support the hypothesis that differences in SCFA concentrations are associated with colorectal cancer.

@@

**Study design and sampling.** The overall protocol has been described in detail previously [@normalization_sze2017; @crc_model_baxter2016]. In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before (pre-) and after (post-) treatment of a patient's tumor (n=`r length(metaf$initial)`). For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from `r min(metaf$time)` - `r max(metaf$time)` days. Our use of treatment has been previously defined as encompassing removal of a tumor with or without chemotherapy and radiation [@normalization_sze2017]. Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@normalization_sze2017; @crc_model_baxter2016]. The University of Michigan Institutional Review Board approved the study and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Declaration.

**Measuring specific SCFAs.** Our protocol for the measurement of acetate, butyrate, and propionate followed a previously published protocol that used a High-Performance Liquid Chromatography (HPLC) machine [@scfa_measures_venkataraman2016]. The following changes to this protocol included the use of frozen fecal samples suspended in 1ml of PBS instead of fecal suspensions in DNA Genotek OmniGut tubes, and the use of the actual weight of fecal samples instead of the average weight for SCFA concentration normalizations. These methodological changes did not affect the overall median concentrations of these SCFAs between the two studies (see Table 1 [@scfa_measures_venkataraman2016] and Figure 1 here).

@@

**Generating imputed metagenomes.** The use of PICRUSt version 1.1.2 with the recommended standard operating protocol [@Langille2013] was used. Briefly, the mothur shared file and metadata was converted into a biom formatted table using the biom convert function, the subsequent biom file was processed with the 'normalize_by_copy_number.py' function, and subsequent imputed metagenomes created using the 'predict_metagenomes.py' function.

@@

The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study. We would also like to thank Kwi Kim and Thomas M Schmidt for their help in running the short-chain fatty acid analysis on the High-Performance Liquid Chromatography machine at the University of Michigan. Salary support for Marc A. Sze came from the Canadian Institute of Health Research and NIH grant UL1TR002240. Salary support for Patrick D. Schloss came from NIH grants P30DK034933 and 1R01CA215574.

@@

**Figure S1. OTU-based Random Forest models of SCFA concentrations.** The train and test results of 100 different 80/20 OTU-based models with 10-fold CV based on higher or lower than the median SCFA concentration using classification Random Forest (A). The top 10 OTUs based on mean decrease in accuracy (MDA) for each model, colored by their lowest taxonomic identification (B).

### 2018-06-14 submission/manuscript.tex
spell check and add nick changes to abstract

\textbf{Background.} Colorectal cancer (CRC) is a growing health concern
with a large percentage of the risk of developing disease being due to
environmental factors. Certain bacterial community members have been
associated to increase CRC while other taxa have been associated with
normal colons. Some of the taxa associated with normal colons can use
fiber to produce short-chain fatty acids (SCFAs) that can inhibit tumor
growth in model systems. However, the data supporting the importance of
SCFAs in human CRC is less certain. Here, we test the hypothesis that
SCFA concentration changes in individuals with colorectal tumors.

\textbf{Methods.} We analyzed a cross-sectional (n=490) and pre- and
post-treatment (n=67) group for their concentrations of acetate,
butyrate, and propionate. Analysis also included tumor classification
models using Random Forest, imputed gene relative abundance with
PICRUSt, and metagenomic sequencing on a subset (n=85) of the total
cross-sectional group.

\textbf{Results.} No difference in SCFA concentration was found between
normal individuals and patients with adenoma or carcinoma (P-value
\textgreater{} 0.15). There was no difference in classification models
with or without SCFAs in their ability to predict adenoma or carcinoma
(P-value \textgreater{} 0.05). Using metagenomic sequencing, there was
also no difference in genes involved with SCFA synthesis (P-value
\textgreater{} 0.70).

\textbf{Conclusions.} Although our data does not support the hypothesis
that SCFAs decrease in individuals that have colorectal tumors, there
may be specific scenarios where SCFAs may still be beneficial for
treatment of CRC. Alternatively, there may be other metabolites that
have not been thoroughly investigated that are more important to the
development of human CRC.

@@

Colorectal cancer (CRC) is currently the third leading cancer-related
cause of death within the US (1, 2). Although there is a genetic
component to the disease, the envrionment has been attributed as an
important and large risk factor for CRC (3). This has lead to the
investigation of how the microbiota may exacerbate or cause tumorigensis
(4--6) and whether the bacterial community is altered (7, 8). Multiple
reports in case/control studies on the microbiota and CRC have
identified taxa commonly associated with the normal colons to be
decreased in patients with carcinoma tumors (9--11). Many of these taxa
actively produce short-chain fatty acids (SCFAs) from fiber that are a
part of our general diet (12). The most extensively studied of these
SCFAs are acetate, butyrate, and propionate (13). Overall, CRC is a
growing health problem and the microbiota through SCFAs have shown
promise as a potential treatment option.

@@

To accomplish this task we will directly measure the concentration of
acetate, butyrate, and propionate within fecal samples for two different
groups. The first group had a sample obtained at a single cross
sectional point in time while the second group had samples obtained
before (pre-) and after (post-)treatment for colorectal tumors.
Additionally, we also assessed whether adding SCFA concentrations to OTU
data improves the classification of adenoma or carcinoma using the
Random Forest algorithm (19), used PICRUSt (20) and metagenomic
sequencing to assess the genes involved in SCFA synthesis, and analyzed
how well 16S rRNA gene sequencing can predict SCFA concentrations. This
investigation provides additional information as to whether the taxa
that are decreased in carcionma use SCFAs to help protect against colon
tumorigenesis.

@@

\textbf{Changes in genes for enzymes involved in SCFA synthesis are not
associated with tumors.} Using a list of specific genes that are
important for the synthesis of SCFAs {[}Table S1{]}, we looked for
differences in gene abundance based on individuals having normal colons,
adenomas, or carcinomas. First, using imputed gene relative abundance
based on 16S rRNA gene sequencing we found no difference in any of the
genes inolved with acetate, butyrate, or propionate synthesis (P-value
\textgreater{} 0.90) {[}Table S2{]}. This similarity between groups is
highlighted by visualizing genes important in butyrate synthesis
{[}Figure 3A{]}. Using a Wilcoxson rank-sum test, there also was no
difference in imputed gene relative abundance between pre- and
post-treatment samples for any genes involved with SCFA synthesis
(P-value \textgreater{} 0.70) {[}Table S3{]}. Next, we took a subset of
these 490 fecal samples (n=85) and used metagenomic sequencing to
confirm these results. Like the imputed gene results, metagenomic
analysis found that there was no difference in any of the genes involved
in SCFA synthesis between individuals with normal colons (n=29) and
patients with adenoma (n=28) or carcinoma (n=28) (P-value \textgreater{}
0.70) {[}Table S4{]}. This lack of difference is highlighted when we
visualize the results for butyrate kinase {[}Figure 3B{]}. These
observations provide evidence that the gene content does not change due
to colorectal tumors. Although these results suggest that SCFA
concentrations do not change due to colorectal tumor, there may be some
errors present within our results that may account for the lack of
associations.

\textbf{Expected taxa are associated with higher SCFA concentrations
regardless of tumor status.} Using OTU data we built Random Forest
models to classify higher than median and lower than median SCFA
concentrations. Overall, OTU data had a reasonable ability to classify
high and low SCFA concentrations {[}Figure S1A{]}. However, these models
tended to be overfit, suggesting that rarer taxa may be important for
this classification {[}Figure S1A{]}. The most important OTUs to these
models (assessed with mean decrease in accuracy (MDA)) were to taxa that
are normally associated with SCFA production {[}Figure S1B{]}. These
results highlight that there are no immediately apparent errors in our
data because SCFA concentrations are associated with taxa known to
produce acetate, butyrate, and propionate. Additionally, OTUs associated
with these taxa are the most important to models that can classify high
and low SCFA concentrations. Overall, our results are robust and do not
support the hypothesis that differences in SCFA concentrations are
associated with colorectal cancer.

@@

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (26, 27). In brief, this study used fecal
samples obtained at either a single cross-sectional time point (n=490)
or from before (pre-) and after (post-) treatment of a patient's tumor
(n=67). For patients undergoing treatment for their tumor the length of
time between their initial and follow up sample ranged from 188 - 546
days. Our use of treatment has been previously defined as encompassing
removal of a tumor with or without chemotherapy and radiation (26).
Diagnosis of tumor was made by colonoscopic examination and
histopathological review of biopsies obtained (26, 27). The University
of Michigan Institutional Review Board approved the study and informed
consent was obtained from all participants in accordance to the
guidelines set out by the Helsinki Decleration.

\textbf{Measuring specific SCFAs.} Our protocol for the measurement of
acetate, butyrate, and propionate followed a previously published
protocol that used a High-Performance Liquid Chromatography (HPLC)
machine (28). The following changes to this protocol included the use of
frozen fecal samples suspended in 1ml of PBS instead of fecal
suspensions in DNA Genotek OmniGut tubes, and the use of the acutal
weight of fecal samples instead of the average weight for SCFA
concentration normalizations. These methodological changes did not
affect the overall median concentrations of these SCFAs between the two
studies (see Table 1 (28) and Figure 1 here).

@@

\textbf{Generating imputed metagenomes.} The use of PICRUSt version
1.1.2 with the recommended standard operating protocol (20) was used.
Briefly, the mothur shared file and metadata was converted into a biom
formated table using the biom convert function, the subsequent biom file
was processed with the `normalize\_by\_copy\_number.py' function, and
subsequent imputed metagenomes created using the
`predict\_metagenomes.py' function.

@@

\textbf{Figure S1. OTU-based Random Forest models of SCFA
concentrations.} The train and test results of 100 different 80/20
OTU-based models with 10-fold CV based on higher or lower than the
medain SCFA concentration using classification Random Forest (A). The
top 10 OTUs based on mean decrease in accuracy (MDA) for each model,
colored by their lowest taxonomic identification (B).

>>>>

\textbf{Background.} Colorectal cancer (CRC) is a growing health concern
with the majority of the risk for developing disease being due to
environmental factors. The microbiota is one of these environmental
factors with certain bacterial community members being associated with
CRC, while other taxa are associated to colons without tumors. Some of
the taxa associated to colons without tumors can use fiber to produce
short-chain fatty acids (SCFAs) that can inhibit tumor growth in model
systems. However, the data supporting the importance of SCFAs in human
CRC is less certain. Here, we test the hypothesis that SCFA
concentrations are different in individuals with colorectal tumors.

\textbf{Methods.} We analyzed a cross-sectional (n=490) and longitudinal
pre- and post-treatment (n=67) group for their concentrations of
acetate, butyrate, and propionate. Analysis also included tumor
classification models using Random Forest, imputed gene relative
abundance with PICRUSt, and metagenomic sequencing on a subset (n=85) of
the total cross-sectional group.

\textbf{Results.} No difference in SCFA concentrations were found
between individuals without tumors and patients with adenomas or
carcinomas (P-value \textgreater{} 0.15). There was no difference in
classification models with or without SCFAs in their ability to predict
patients with adenomas or carcinomas versus individuals without tumors
(P-value \textgreater{} 0.05). Using metagenomic sequencing, there was
also no difference in genes involved with SCFA synthesis between
individuals without tumors and patients with adenomas or carcinomas
(P-value \textgreater{} 0.70).

\textbf{Conclusions.} Although our data does not support the hypothesis
that SCFAs are different in individuals that have colorectal tumors,
there may be context specific scenarios where SCFAs may still be
beneficial for treatment of CRC. Alternatively, there may be other
mechanisms that have not been thoroughly investigated that are more
important to the development of human CRC.

@@

Colorectal cancer (CRC) is currently the third leading cancer-related
cause of death within the US (1, 2). Although there is a genetic
component to the disease, the environment has been attributed as an
important and large risk factor for CRC (3). This has lead to the
investigation of how the microbiota may exacerbate or cause tumorigensis
(4--6) and whether the bacterial community is altered (7, 8). Multiple
reports in case/control studies on the microbiota and CRC have
identified taxa commonly associated with the normal colons to be
decreased in patients with carcinoma tumors (9--11). Many of these taxa
actively produce short-chain fatty acids (SCFAs) from fiber that are a
part of our general diet (12). The most extensively studied of these
SCFAs are acetate, butyrate, and propionate (13). Overall, CRC is a
growing health problem and the microbiota through SCFAs have shown
promise as a potential treatment option.

@@

To accomplish this task we will directly measure the concentration of
acetate, butyrate, and propionate within fecal samples for two different
groups. The first group had a sample obtained at a single cross
sectional point in time while the second group had samples obtained
before (pre-) and after (post-)treatment for colorectal tumors.
Additionally, we also assessed whether adding SCFA concentrations to OTU
data improves the classification of adenoma or carcinoma using the
Random Forest algorithm (19), used PICRUSt (20) and metagenomic
sequencing to assess the genes involved in SCFA synthesis, and analyzed
how well 16S rRNA gene sequencing can predict SCFA concentrations. This
investigation provides additional information as to whether the taxa
that are decreased in carcinoma use SCFAs to help protect against colon
tumorigenesis.

@@

\textbf{Changes in genes for enzymes involved in SCFA synthesis are not
associated with tumors.} Using a list of specific genes that are
important for the synthesis of SCFAs {[}Table S1{]}, we looked for
differences in gene abundance based on individuals having normal colons,
adenomas, or carcinomas. First, using imputed gene relative abundance
based on 16S rRNA gene sequencing we found no difference in any of the
genes involved with acetate, butyrate, or propionate synthesis (P-value
\textgreater{} 0.90) {[}Table S2{]}. This similarity between groups is
highlighted by visualizing genes important in butyrate synthesis
{[}Figure 3A{]}. Using a Wilcoxon rank-sum test, there also was no
difference in imputed gene relative abundance between pre- and
post-treatment samples for any genes involved with SCFA synthesis
(P-value \textgreater{} 0.70) {[}Table S3{]}. Next, we took a subset of
these 490 fecal samples (n=85) and used metagenomic sequencing to
confirm these results. Like the imputed gene results, metagenomic
analysis found that there was no difference in any of the genes involved
in SCFA synthesis between individuals with normal colons (n=29) and
patients with adenoma (n=28) or carcinoma (n=28) (P-value \textgreater{}
0.70) {[}Table S4{]}. This lack of difference is highlighted when we
visualize the results for butyrate kinase {[}Figure 3B{]}. These
observations provide evidence that the gene content does not change due
to colorectal tumors. Although these results suggest that SCFA
concentrations do not change due to colorectal tumor, there may be some
errors present within our results that may account for the lack of
associations.

\textbf{Expected taxa are associated with higher SCFA concentrations
regardless of tumor status.} Using OTU data we built Random Forest
models to classify higher than median and lower than median SCFA
concentrations. Overall, OTU data had a reasonable ability to classify
high and low SCFA concentrations {[}Figure S1A{]}. However, these models
tended to be over fit, suggesting that rarer taxa may be important for
this classification {[}Figure S1A{]}. The most important OTUs to these
models (assessed with mean decrease in accuracy (MDA)) were to taxa that
are normally associated with SCFA production {[}Figure S1B{]}. These
results highlight that there are no immediately apparent errors in our
data because SCFA concentrations are associated with taxa known to
produce acetate, butyrate, and propionate. Additionally, OTUs associated
with these taxa are the most important to models that can classify high
and low SCFA concentrations. Overall, our results are robust and do not
support the hypothesis that differences in SCFA concentrations are
associated with colorectal cancer.

@@

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (26, 27). In brief, this study used fecal
samples obtained at either a single cross-sectional time point (n=490)
or from before (pre-) and after (post-) treatment of a patient's tumor
(n=67). For patients undergoing treatment for their tumor the length of
time between their initial and follow up sample ranged from 188 - 546
days. Our use of treatment has been previously defined as encompassing
removal of a tumor with or without chemotherapy and radiation (26).
Diagnosis of tumor was made by colonoscopic examination and
histopathological review of biopsies obtained (26, 27). The University
of Michigan Institutional Review Board approved the study and informed
consent was obtained from all participants in accordance to the
guidelines set out by the Helsinki Declaration.

\textbf{Measuring specific SCFAs.} Our protocol for the measurement of
acetate, butyrate, and propionate followed a previously published
protocol that used a High-Performance Liquid Chromatography (HPLC)
machine (28). The following changes to this protocol included the use of
frozen fecal samples suspended in 1ml of PBS instead of fecal
suspensions in DNA Genotek OmniGut tubes, and the use of the actual
weight of fecal samples instead of the average weight for SCFA
concentration normalizations. These methodological changes did not
affect the overall median concentrations of these SCFAs between the two
studies (see Table 1 (28) and Figure 1 here).

@@

\textbf{Generating imputed metagenomes.} The use of PICRUSt version
1.1.2 with the recommended standard operating protocol (20) was used.
Briefly, the mothur shared file and metadata was converted into a biom
formatted table using the biom convert function, the subsequent biom
file was processed with the `normalize\_by\_copy\_number.py' function,
and subsequent imputed metagenomes created using the
`predict\_metagenomes.py' function.

@@

\textbf{Figure S1. OTU-based Random Forest models of SCFA
concentrations.} The train and test results of 100 different 80/20
OTU-based models with 10-fold CV based on higher or lower than the
median SCFA concentration using classification Random Forest (A). The
top 10 OTUs based on mean decrease in accuracy (MDA) for each model,
colored by their lowest taxonomic identification (B).
